---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NPLStoolbox

<!-- badges: start -->
[![R-CMD-check](https://github.com/GRvanderPloeg/NPLStoolbox/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/GRvanderPloeg/NPLStoolbox/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/GRvanderPloeg/NPLStoolbox/graph/badge.svg)](https://app.codecov.io/gh/GRvanderPloeg/NPLStoolbox)
<!-- badges: end -->

The `NPLStoolbox` allows researchers to use the N-way Partial Least Squares method for their multi-way data.

Currently it contains a work in process `triPLS1()` function and an example dataset `Cornejo2025`.

More will follow!

## Installation

You can install the development version of NPLStoolbox from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("GRvanderPloeg/NPLStoolbox")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(NPLStoolbox)
## basic example code
#TBD
```
