---
title: "MAINHEALTH: intergenerational obesity transmission"
output: html_document
date: "2025-08-31"
---

# Introduction

This article presents the NPLS results of the MAINHEALTH dataset discussed in Chapter 5 of my dissertation.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preamble

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(parafac4microbiome)
library(NPLStoolbox)
library(CMTFtoolbox)
library(ggplot2)
library(ggpubr)
library(stringr)
library(scales)
library(ggpattern)
```

```{r metadata test function}
testMetadata = function(model, comp, metadata){
  transformedSubjectLoadings = model$Fac[[1]][,comp]
  transformedSubjectLoadings = transformedSubjectLoadings / norm(transformedSubjectLoadings, "2")
  
  result = lm(transformedSubjectLoadings ~ BMI + C.section + Secretor + Lewis + whz.6m, data=metadata$mode1)
  
  # Extract coefficients and confidence intervals
  coef_estimates <- summary(result)$coefficients
  conf_intervals <- confint(result)
  
  # Remove intercept
  coef_estimates <- coef_estimates[rownames(coef_estimates) != "(Intercept)", ]
  conf_intervals <- conf_intervals[rownames(conf_intervals) != "(Intercept)", ]
  
  # Combine into a clean data frame
  summary_table <- data.frame(
    Term     = rownames(coef_estimates),
    Estimate = coef_estimates[, "Estimate"] * 1e3,
    CI       = paste0(
                 conf_intervals[, 1], " – ",
                 conf_intervals[, 2]
               ),
    P_value  = coef_estimates[, "Pr(>|t|)"],
    P_adjust = p.adjust(coef_estimates[, "Pr(>|t|)"], "BH"),
    row.names = NULL
  )
  
  return(summary_table)
}
```

```{r define colours}
colours = RColorBrewer:: brewer.pal(8, "Dark2")
BMI_cols = c("darkgreen","darkgoldenrod","darkred") #colours[1:3]
WHZ_cols = c("darkgreen", "darkgoldenrod", "dodgerblue4")
phylum_cols = colours
```

# Maternal pre-pregnancy BMI

## Processing

```{r process data}
# Faeces
newFaeces = NPLStoolbox::Jakobsen2025$faeces

mask = newFaeces$mode1$subject %in% (NPLStoolbox::Jakobsen2025$faeces$mode1 %>% filter(!is.na(BMI)) %>% select(subject) %>% pull())
newFaeces$data = newFaeces$data[mask,,]
newFaeces$mode1 = newFaeces$mode1[mask,]
processedFaeces = parafac4microbiome::processDataCube(newFaeces, sparsityThreshold = 0.75, centerMode=1, scaleMode=2)

# Milk
newMilk = NPLStoolbox::Jakobsen2025$milkMicrobiome

mask = newMilk$mode1$subject %in% (NPLStoolbox::Jakobsen2025$milkMicrobiome$mode1 %>% filter(!is.na(BMI)) %>% select(subject) %>% pull())
newMilk$data = newMilk$data[mask,,]
newMilk$mode1 = newMilk$mode1[mask,]
processedMilk = parafac4microbiome::processDataCube(newMilk, sparsityThreshold=0.85, centerMode=1, scaleMode=2)

# Milk metab
newMilkMetab = NPLStoolbox::Jakobsen2025$milkMetabolomics

mask = newMilkMetab$mode1$subject %in% (NPLStoolbox::Jakobsen2025$milkMetabolomics$mode1 %>% filter(!is.na(BMI)) %>% select(subject) %>% pull())
newMilkMetab$data = newMilkMetab$data[mask,,]
newMilkMetab$mode1 = newMilkMetab$mode1[mask,]
processedMilkMetab = parafac4microbiome::processDataCube(newMilkMetab, CLR=FALSE, centerMode=1, scaleMode=2)
```

```{r prepare y}
Y = processedFaeces$mode1$BMI
Ycnt = Y - mean(Y)
Ynorm_faecal = Ycnt / norm(Ycnt, "2")

Y = processedMilk$mode1$BMI
Ycnt = Y - mean(Y)
Ynorm_milk = Ycnt / norm(Ycnt, "2")

Y = processedMilkMetab$mode1$BMI
Ycnt = Y - mean(Y)
Ynorm_milkMetab = Ycnt / norm(Ycnt, "2")
```

## Infant faecal microbiome

NPLS was next applied to supervise the decomposition using maternal pre-pregnancy BMI (ppBMI) as dependent variable (y). The model selection procedure revealed that a one-component NPLS model corresponded to the RMSECV minimum, explaining 5.8% of the variation in X, and 12.0% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r faecal bmi model selection}
# CV_faecal = NPLStoolbox::ncrossreg(processedFaeces$data, Ynorm_faecal)
# a=CV_faecal$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_faecal$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 comp
```

```{r faecal bmi load model}
# NPLS_faecal = NPLStoolbox::triPLS1(processedFaeces$data, Ynorm_faecal, 1)
NPLS_faecal = readRDS("./MAINHEALTH/20250702_NPLS_faecal_bmi.RDS")

NPLS_faecal$varExpX
NPLS_faecal$varExpY
```

As expected, MLR analysis revealed that the model captured variation exclusively associated with maternal ppBMI.

```{r faecal bmi check metadata}
testMetadata(NPLS_faecal, 1, processedFaeces)
```

```{r npls faecal comp 1}
# subject mode
a = processedFaeces$mode1 %>%
  mutate(comp1 = NPLS_faecal$Fac[[1]][,1]) %>%
  ggplot(aes(x=BMI,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Maternal ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedFaeces$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_faecal$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedFaeces$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_faecal$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_faecal$Fac[[1]][,1], NPLS_faecal$Fac[[2]][,1], NPLS_faecal$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedFaeces$mode1$BMI))
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedFaeces$mode1$BMI))

df = processedFaeces$mode2 %>%
  mutate(Component_1 = NPLS_faecal$Fac[[2]][,1]) %>%
  arrange(Component_1) %>%
  mutate(index=1:nrow(.)) %>%
  mutate(Genus = gsub("g__", "", V7), Species = gsub("s__", "", V8)) %>% 
  mutate(Species = str_split_fixed(Species, "_", 2)[,2]) %>% 
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "sp."), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "organism"), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "bacterium"), "", .x))) %>%
  filter(index %in% c(1:10, 84:93))

df[df$Genus == "" & df$Species == "", "Species"] = "Unknown"
df[df$Species == "", "Species"] = "sp."
df$name = paste0(df$Genus, " ", df$Species)
  
b = df %>% ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(V3))) +
  geom_bar(stat="identity", col="black") +
  xlab("Loading") +
  ylab("") +
  scale_y_discrete(labels=df$name) +
  scale_fill_manual(name="Phylum", values=hue_pal()(5)[-5], labels=c("Actinobacteriota", "Bacteroidota", "Firmicutes", "Proteobacteria")) +
  theme(text=element_text(size=16))

# time mode
c = NPLS_faecal$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=value) %>% 
  ggplot(aes(x=c(30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

Positively loaded microbiota including Bifidobacterium breve, Bifidobacterium bifidum, Enterococcus sp., Escherichia coli, Clostridium} sp., and Staphylococcus epidermidis were enriched in infants of high ppBMI mothers and depleted in infants of low ppBMI mothers. Conversely, negatively loaded microbiota including Libanicoccus, Streptococcus, Actinomyces, and Klebsiella sp., as well as Citrobacter freundii, Rothia mucilaginosa, Enterococcus faecalis, Bifidobacterium adolescentis, and Parabacteroides distasonis were enriched in infants of low ppBMI mothers and depleted in infants of high ppBMI mothers. The time mode loadings indicate that ppBMI-associated inter-subject differences peaked at day 60. 

# HM microbiome

NPLS was next applied to supervise the decomposition using maternal pre-pregnancy BMI (ppBMI) as dependent variable (y). The model selection procedure revealed that a one-component model corresponded to the RMSECV minimum, explaining 5.4% of the variation in X, and 11.2% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm microbiome bmi model selection}
# CV_milk = NPLStoolbox::ncrossreg(processedMilk$data, Ynorm_milk)
# a=CV_milk$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_milk$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 or 2 comps
```

```{r hm microbiome bmi load model}
# NPLS_milk = NPLStoolbox::triPLS1(processedMilk$data, Ynorm_milk, 1)
NPLS_milk = readRDS("./MAINHEALTH/20250702_NPLS_milk_bmi.RDS")

NPLS_milk$varExpX
NPLS_milk$varExpY
```

As expected, MLR analysis revealed that the model captured variation exclusively associated with maternal ppBMI.

```{r hm microbiome bmi check metadata}
testMetadata(NPLS_milk, 1, processedMilk)
```

```{r npls milk comp 1}
# subject mode
a = processedMilk$mode1 %>%
  mutate(comp1 = NPLS_milk$Fac[[1]][,1]) %>%
  ggplot(aes(x=BMI,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Maternal ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milk$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milk$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_milk$Fac[[1]][,1], NPLS_milk$Fac[[2]][,1], NPLS_milk$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilk$mode1$BMI)) # flip
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilk$mode1$BMI))

df = processedMilk$mode2 %>%
  mutate(Component_1 = -1*NPLS_milk$Fac[[2]][,1]) %>%
  arrange(Component_1) %>%
  mutate(index=1:nrow(.)) %>%
  mutate(Genus = gsub("g__", "", V7), Species = gsub("s__", "", V8)) %>% 
  mutate(Species = str_split_fixed(Species, "_", 2)[,2]) %>% 
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "sp."), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "organism"), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "bacterium"), "", .x))) %>%
  filter(index %in% c(1:10, 106:115))

df[df$Genus == "" & df$Species == "", "Species"] = "Unknown"
df[df$Species == "", "Species"] = "sp."
df$name = paste0(df$Genus, " ", df$Species)
  
b = df %>% ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(V3))) +
  geom_bar(stat="identity", col="black") +
  xlab("Loading") +
  ylab("") +
  scale_y_discrete(labels=df$name) +
  scale_fill_manual(name="Phylum", values=hue_pal()(5)[-5], labels=c("Actinobacteriota", "Bacteroidota", "Firmicutes", "Proteobacteria")) +
  theme(text=element_text(size=16))

# time mode
c = NPLS_milk$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=-1*value) %>% 
  ggplot(aes(x=c(0,30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

Positively loaded microbiota including Gemella, Streptococcus and Staphylococcus spp. were enriched in high ppBMI mothers and depleted in low ppBMI mothers. Conversely, negatively loaded microbiota including Ligilactobacillus murinus, Clostridium sensu stricto spp., Cutibacterium acnes, Terrisporobacter sp., Lactobacillus gasseri, and Stenotrophomonas maltophilia were enriched in low ppBMI mothers and depleted in high ppBMI mothers. The time mode loadings indicated that inter-subject differences due to maternal ppBMI were largest at day 0 and declined over time. 

# HM metabolome

NPLS was next applied to supervise the decomposition using maternal pre-pregnancy BMI (ppBMI) as dependent variable (y). While the model selection procedure revealed that a two-component NPLS model corresponded to the RMSECV minimum, we found that the extra component did not help in the prediction of y, nor was it biologically interpretable. Hence, a one-component NPLS model was fitted, explaining 3.1% of the variation in X, and 31.0% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm metabolome bmi model selection}
# CV_milkMetab = NPLStoolbox::ncrossreg(processedMilkMetab$data, Ynorm_milkMetab)
# a=CV_milkMetab$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_milkMetab$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 comp
```

```{r hm metabolome bmi load model}
# NPLS_milkMetab = NPLStoolbox::triPLS1(processedMilkMetab$data, Ynorm_milkMetab, 2)
NPLS_milkMetab = readRDS("./MAINHEALTH/20250702_NPLS_milkMetab_bmi.RDS")

NPLS_milkMetab$varExpX
NPLS_milkMetab$varExpY
```

However, MLR analysis revealed that the model captured variation associated with a mixture of maternal ppBMI, maternal secretor status, and infant growth at 6 months.

```{r hm metabolome bmi check metadata}
testMetadata(NPLS_milkMetab, 1, processedMilkMetab)
```

```{r npls milkMetab comp 1}
# subject mode
a = processedMilkMetab$mode1 %>%
  mutate(comp1 = NPLS_milkMetab$Fac[[1]][,1]) %>%
  ggplot(aes(x=BMI,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Maternal ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_milkMetab$Fac[[1]][,1], NPLS_milkMetab$Fac[[2]][,1], NPLS_milkMetab$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilkMetab$mode1$BMI))
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilkMetab$mode1$BMI))

temp = processedMilkMetab$mode2 %>% mutate(Component_1 = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(Component_1) %>% mutate(index=1:nrow(.)) %>% filter(index %in% c(1:10, 61:70))

b=temp %>%
  ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(Class),pattern=as.factor(Class))) +
  geom_bar_pattern(stat="identity",
                   colour="black",
                   pattern="stripe",
                   pattern_fill="black",
                   pattern_density=0.2,
                   pattern_spacing=0.05,
                   pattern_angle=45,
                   pattern_size=0.2) +
  scale_y_discrete(label=temp$Metabolite) +
  xlab("Loading") +
  ylab("") +
  scale_fill_manual(name="Class", values=hue_pal()(7)[-5], labels=c("Amino acids and derivatives", "Energy related", "Fatty acids and derivatives", "Food derived", "Oligosaccharides", "Sugars")) +
  theme(text=element_text(size=16), legend.position="none")

# time mode
c = NPLS_milkMetab$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=value) %>% 
  ggplot(aes(x=c(0,30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

# Infant weight-over-height z-score at six months

## Processing

```{r process and homogenize data whz}
# Faeces
newFaeces = NPLStoolbox::Jakobsen2025$faeces

mask1 = newFaeces$mode1$subject %in% (NPLStoolbox::Jakobsen2025$faeces$mode1 %>% filter(!is.na(whz.6m)) %>% select(subject) %>% pull())
mask2 = newFaeces$mode1$subject != "322"
mask = mask1 & mask2
newFaeces$data = newFaeces$data[mask,,]
newFaeces$mode1 = newFaeces$mode1[mask,]
processedFaeces = parafac4microbiome::processDataCube(newFaeces, sparsityThreshold = 0.75, centerMode=1, scaleMode=2) # was 0.75

# Milk
newMilk = NPLStoolbox::Jakobsen2025$milkMicrobiome

mask = newMilk$mode1$subject %in% (NPLStoolbox::Jakobsen2025$milkMicrobiome$mode1 %>% filter(!is.na(whz.6m)) %>% select(subject) %>% pull())
newMilk$data = newMilk$data[mask,,]
newMilk$mode1 = newMilk$mode1[mask,]
processedMilk = parafac4microbiome::processDataCube(newMilk, sparsityThreshold=0.85, centerMode=1, scaleMode=2) # was 0.85

# Milk metab
newMilkMetab = NPLStoolbox::Jakobsen2025$milkMetabolomics

mask = newMilkMetab$mode1$subject %in% (NPLStoolbox::Jakobsen2025$milkMetabolomics$mode1 %>% filter(!is.na(whz.6m)) %>% select(subject) %>% pull())
newMilkMetab$data = newMilkMetab$data[mask,,-1]
newMilkMetab$mode1 = newMilkMetab$mode1[mask,]
newMilkMetab$mode3 = newMilkMetab$mode3[-1,]
processedMilkMetab = parafac4microbiome::processDataCube(newMilkMetab, CLR=FALSE, centerMode=1, scaleMode=2)
```

```{r prepare y whz}
Y = processedFaeces$mode1$whz.6m
Ycnt = Y - mean(Y)
Ynorm_faecal = Ycnt / norm(Ycnt, "2")

Y = processedMilk$mode1$whz.6m
Ycnt = Y - mean(Y)
Ynorm_milk = Ycnt / norm(Ycnt, "2")

Y = processedMilkMetab$mode1$whz.6m
Ycnt = Y - mean(Y)
Ynorm_milkMetab = Ycnt / norm(Ycnt, "2")
```

## Infant faecal microbiome

Subsequently, NPLS was applied to supervise the decomposition using the infant growth weight-for-height at six months (WHZ) as dependent variable (y). The model selection procedure revealed that a one-component NPLS model corresponded to the RMSECV minimum, explaining 2.6% of the variation in X, and 11.2% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r faecal whz model selection}
# CV_faecal = NPLStoolbox::ncrossreg(processedFaeces$data, Ynorm_faecal)
# a=CV_faecal$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_faecal$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 comp
```

```{r faecal whz load model}
# NPLS_faecal = NPLStoolbox::triPLS1(processedFaeces$data, Ynorm_faecal, 1)
NPLS_faecal = readRDS("./MAINHEALTH/20250702_NPLS_faecal_whz.RDS")

NPLS_faecal$varExpX
NPLS_faecal$varExpY
```

As expected, MLR analysis revealed that the model captured variation exclusively associated with maternal ppBMI.

```{r faecal metadata test whz}
testMetadata(NPLS_faecal, 1, processedFaeces)
```

```{r npls faecal comp 1 whz}
# subject mode
a = processedFaeces$mode1 %>%
  mutate(comp1 = NPLS_faecal$Fac[[1]][,1]) %>%
  ggplot(aes(x=whz.6m,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Infant 6-month WHZ") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedFaeces$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_faecal$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedFaeces$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_faecal$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_faecal$Fac[[1]][,1], NPLS_faecal$Fac[[2]][,1], NPLS_faecal$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedFaeces$mode1$whz.6m))
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedFaeces$mode1$whz.6m))

df = processedFaeces$mode2 %>%
  mutate(Component_1 = -1*NPLS_faecal$Fac[[2]][,1]) %>%
  arrange(Component_1) %>%
  mutate(index=1:nrow(.)) %>%
  mutate(Genus = gsub("g__", "", V7), Species = gsub("s__", "", V8)) %>% 
  mutate(Species = str_split_fixed(Species, "_", 2)[,2]) %>% 
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "sp."), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "organism"), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "bacterium"), "", .x))) %>%
  filter(index %in% c(1:10, 84:93))

df[df$Genus == "" & df$Species == "", "Species"] = "Unknown"
df[df$Species == "", "Species"] = "sp."
df$name = paste0(df$Genus, " ", df$Species)
  
b = df %>% ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(V3))) +
  geom_bar(stat="identity", col="black") +
  xlab("Loading") +
  ylab("") +
  scale_y_discrete(labels=df$name) +
  scale_fill_manual(name="Phylum", values=hue_pal()(5)[-5], labels=c("Actinobacteriota", "Bacteroidota", "Firmicutes", "Proteobacteria")) +
  theme(text=element_text(size=16))

# time mode
c = NPLS_faecal$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=-1*value) %>% 
  ggplot(aes(x=c(30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

Positively loaded microbiota included many Actinobacteriota such as Actinomyces sp., Bifidobacterium bifidum, Bifidobacterium breve, and Bifidobacterium longum}, as well as Enterococcus sp., Straphylococcus aureus, and Lacticaseibacillus rhamnosus, which were enriched in high-growth infants and depleted in low-growth infants. Conversely, negatively loaded microbiota included many Proteobacteria such as Escherichia-Shigella coli, Klepsiella spp., and Haemophilus sp., as well as Erysipelatoclostridium spp., Parabacteroides sp., Bacteroides fragilis, and Clostridium sensu stricto sp., which were enriched in low-growth infants and depleted in high-growth infants. The time mode loadings indicated that these inter-subject differences were maximal at day 90.

## HM microbiome

Subsequently, NPLS was applied to supervise the decomposition using the infant growth weight-for-height at six months (WHZ) as dependent variable (y). While the model selection procedure revealed that a two-component model corresponded to the RMSECV minimum, we found that the extra component did not help in the prediction of y, nor was it biologically interpretable. Hence, a one-component NPLS model was fitted, explaining 1.6% of the variation in X, and 9.7% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm microbiome whz model selection}
# CV_milk = NPLStoolbox::ncrossreg(processedMilk$data, Ynorm_milk)
# a=CV_milk$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_milk$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 or 2 comps
```

```{r hm microbiome whz load model}
# NPLS_milk = NPLStoolbox::triPLS1(processedMilk$data, Ynorm_milk, 1)
NPLS_milk = readRDS("./MAINHEALTH/20250702_NPLS_milk_whz.RDS")

NPLS_milk$varExpX
NPLS_milk$varExpY
```

However, MLR analysis revealed that the model captured variation associated with a mixture of infant WHZ and maternal ppBMI.

```{r hm microbiome metadata test whz}
testMetadata(NPLS_milk, 1, processedMilk)
```

```{r npls milk comp 1 whz}
# subject mode
a = processedMilk$mode1 %>%
  mutate(comp1 = NPLS_milk$Fac[[1]][,1]) %>%
  ggplot(aes(x=whz.6m,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Maternal ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milk$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilk$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milk$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_milk$Fac[[1]][,1], NPLS_milk$Fac[[2]][,1], NPLS_milk$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilk$mode1$whz.6m)) # flip
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilk$mode1$whz.6m))

df = processedMilk$mode2 %>%
  mutate(Component_1 = -1*NPLS_milk$Fac[[2]][,1]) %>%
  arrange(Component_1) %>%
  mutate(index=1:nrow(.)) %>%
  mutate(Genus = gsub("g__", "", V7), Species = gsub("s__", "", V8)) %>% 
  mutate(Species = str_split_fixed(Species, "_", 2)[,2]) %>% 
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "sp."), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "organism"), "", .x))) %>%
  mutate(dplyr::across(.cols = Species, .fns = ~ dplyr::if_else(stringr::str_detect(.x, "bacterium"), "", .x))) %>%
  filter(index %in% c(1:10, 106:115))

df[df$Genus == "" & df$Species == "", "Species"] = "Unknown"
df[df$Species == "", "Species"] = "sp."
df$name = paste0(df$Genus, " ", df$Species)
  
b = df %>% ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(V3))) +
  geom_bar(stat="identity", col="black") +
  xlab("Loading") +
  ylab("") +
  scale_y_discrete(labels=df$name) +
  theme(text=element_text(size=16))

# time mode
c = NPLS_milk$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=-1*value) %>% 
  ggplot(aes(x=c(0,30,60,90),y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```

## HM metabolome

Subsequently, NPLS was applied to supervise the decomposition using the infant growth weight-for-height at six months (WHZ) as dependent variable (y). The model selection procedure revealed that a one-component NPLS model corresponded to the RMSECV minimum, explaining 4.2% of the variation in X, and 15.8% of the variation in y. The commented code below performs the procedure, and is not rerun here due to computational limitations.

```{r hm metabolome whz model selection}
# CV_milkMetab = NPLStoolbox::ncrossreg(processedMilkMetab$data, Ynorm_milkMetab)
# a=CV_milkMetab$varExp %>% pivot_longer(-numComponents) %>% ggplot(aes(x=as.factor(numComponents),y=value,col=as.factor(name),group=as.factor(name))) + geom_line() + geom_point() + xlab("Number of components") + ylab("Variance explained (%)") + theme(legend.position="none")
# b=CV_milkMetab$RMSE %>% ggplot(aes(x=numComponents,y=RMSE)) + geom_line() + geom_point() + xlab("Number of components") + ylab("RMSECV")
# ggarrange(a,b) # 1 comp
```

```{r hm metabolome whz load model}
# NPLS_milkMetab = NPLStoolbox::triPLS1(processedMilkMetab$data, Ynorm_milkMetab, 2)
NPLS_milkMetab = readRDS("./MAINHEALTH/20250702_NPLS_milkMetab_whz.RDS")

NPLS_milkMetab$varExpX
NPLS_milkMetab$varExpY
```

However, MLR analysis revealed that the model captured variation associated with a mixture of infant growth, but also maternal ppBMI and maternal secretor status.

```{r hm metabolome metadata test whz}
testMetadata(NPLS_milkMetab, 1, processedMilkMetab)
```

```{r npls milkMetab comp 1 whz}
# subject mode
a = processedMilkMetab$mode1 %>%
  mutate(comp1 = NPLS_milkMetab$Fac[[1]][,1]) %>%
  ggplot(aes(x=whz.6m,y=comp1)) +
  geom_point() +
  stat_cor() +
  xlab("Maternal ppBMI") +
  ylab("Loading") +
  theme(text=element_text(size=16))

# feature mode
topIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% head() %>% select(index) %>% pull()
bottomIndices = processedMilkMetab$mode2 %>% mutate(index=1:nrow(.), Comp = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(desc(Comp)) %>% tail() %>% select(index) %>% pull()

Xhat = parafac4microbiome::reinflateTensor(NPLS_milkMetab$Fac[[1]][,1], NPLS_milkMetab$Fac[[2]][,1], NPLS_milkMetab$Fac[[3]][,1])

print("Positive loadings:")
print(cor(Xhat[,topIndices[1],1], processedMilkMetab$mode1$whz.6m))
print("Negative loadings:")
print(cor(Xhat[,bottomIndices[1],1], processedMilkMetab$mode1$whz.6m))

temp = processedMilkMetab$mode2 %>% mutate(Component_1 = NPLS_milkMetab$Fac[[2]][,1]) %>% arrange(Component_1) %>% mutate(index=1:nrow(.)) %>% filter(index %in% c(1:10, 61:70))

b=temp %>%
  ggplot(aes(x=Component_1,y=as.factor(index),fill=as.factor(Class),pattern=as.factor(Class))) +
  geom_bar_pattern(stat="identity",
                   colour="black",
                   pattern="stripe",
                   pattern_fill="black",
                   pattern_density=0.2,
                   pattern_spacing=0.05,
                   pattern_angle=45,
                   pattern_size=0.2) +
  scale_y_discrete(label=temp$Metabolite) +
  xlab("Loading") +
  ylab("") +
  theme(text=element_text(size=16), legend.position="none")

# time mode
c = NPLS_milkMetab$Fac[[3]][,1] %>%
  as_tibble() %>% 
  mutate(value=-1*value, timepoints = c(30,60,90)) %>% 
  ggplot(aes(x=timepoints,y=value)) +
  geom_line() +
  geom_point() +
  xlab("Time point [day]") +
  ylab("Loading") +
  ylim(0,1) +
  theme(text=element_text(size=16))

a
b
c
```
